---
title: "Prototype: LLM-generated semantic-similarity weights"
subtitle: "Making a many-to-many crosswalk weighting judgment explicit and reproducible"
format:
  html:
    embed-resources: true
    toc: true
execute:
  echo: true
---

## What this is

A small prototype for [example_aus-us-occupn#3](https://github.com/cynthiahqy/example_aus-us-occupn/issues/3): can an LLM stand in for the hand-authored "committee judgment" step in the `xmap` vignette, for splitting a many-to-many crosswalk link into weights?

This is **not** a benchmark or a validation exercise — there's no ground truth here, and the vignette's own "committee" weights are themselves illustrative, not a real expert panel's output. The point is narrower: a semantic-similarity judgment about which occupation is "more like" which can be made **explicit and reproducible** — a documented prompt, a specific model, a specific date, producing a specific weight — instead of an unrecorded judgment call.

The two-round prompt framework below is adapted from Ma, Huang & Haensch (2025), ["Can Large Language Models Advance Crosswalks? The Case of Danish Occupation Codes"](https://aclanthology.org/2025.naacl-srw.38/), NAACL 2025 SRW. We only need their **Round 1** (similarity rating) — our subset already has real ABS/BLS-published links, so unlike their setup (searching the full cross-product of two codebooks), we only need to rate the links that already exist, then turn the ratings into split weights.

## Setup

```{r}
#| label: setup
#| message: false
library(ellmer)
library(dplyr)
library(tidyr)
library(readr)
library(purrr)

readRenviron(here::here(".Renviron"))
```

```{r}
#| label: paths
subset_dir <- here::here("data", "subset")
```

## Load the subset crosswalks and definitions

```{r}
#| label: load-data
#| message: false
anzsco_isco <- read_csv(file.path(subset_dir, "anzsco_to_isco08_crosswalk.csv"), show_col_types = FALSE)
soc_isco <- read_csv(file.path(subset_dir, "soc2010_to_isco08_crosswalk.csv"), show_col_types = FALSE)

isco_defs <- read_csv(file.path(subset_dir, "isco08_definitions.csv"), show_col_types = FALSE) |>
  transmute(code = as.character(isco08_code), title, definition)

anzsco_defs <- read_csv(file.path(subset_dir, "anzsco1.2_definitions.csv"), show_col_types = FALSE) |>
  transmute(code = as.character(anzsco_code), title, definition)

soc_defs <- read_csv(file.path(subset_dir, "soc2010_definitions.csv"), show_col_types = FALSE) |>
  transmute(code = soc2010_code, title, definition)
```

## Find the many-to-many links that need a weight

We only need to rate a link where its *source* occupation (the one supplying real count data — ANZSCO for Australia, SOC for the USA) maps to more than one ISCO-08 target. A source with exactly one target needs no judgment call: the full weight is 1.

```{r}
#| label: many-to-many-groups
anzsco_many <- anzsco_isco |>
  filter(!is.na(anzsco_code)) |>
  add_count(anzsco_code, name = "n_targets") |>
  filter(n_targets > 1) |>
  transmute(source_code = as.character(anzsco_code), target_code = as.character(isco08_code))

soc_many <- soc_isco |>
  add_count(soc2010_code, name = "n_targets") |>
  filter(n_targets > 1) |>
  transmute(source_code = as.character(soc2010_code), target_code = as.character(isco08_code))

links_to_rate <- bind_rows(
  anzsco_many |> mutate(source_scheme = "ANZSCO"),
  soc_many |> mutate(source_scheme = "SOC2010")
)

links_to_rate
```

## Round 1 prompt (adapted from Ma, Huang & Haensch 2025)

Same question and 5-point scale as the paper's Round 1, with our own definition text substituted in for `{code1}`/`{code2}`, and a constrained `type_enum` output so the model can only answer A-E.

```{r}
#| label: prompt-template
similarity_type <- type_enum(
  values = c("A", "B", "C", "D", "E"),
  description = paste(
    "Similarity rating: A = Extremely similar, B = Very similar,",
    "C = Moderately similar, D = Slightly similar, E = Not similar at all"
  )
)

build_prompt <- function(title1, def1, title2, def2) {
  glue::glue(
    "How similar do you think these two occupational classification codes are ",
    "in terms of tasks, skills, and responsibilities? Please respond with only ",
    "the option letter out of 'A', 'B', 'C', 'D', or 'E'.\n\n",
    "Code1: {title1} - {def1}\n",
    "Code2: {title2} - {def2}\n\n",
    "Options:\n",
    "A Extremely similar\n",
    "B Very similar\n",
    "C Moderately similar\n",
    "D Slightly similar\n",
    "E Not similar at all"
  )
}
```

## Run Round 1 for every source-target pair that needs a weight

One model, one call per pair — small enough (`r nrow(links_to_rate)` pairs) that every rating can be read directly rather than automated further (no Round 2 needed here).

```{r}
#| label: model-config
llm_model <- "gpt-4o-mini"
run_date <- Sys.Date()
```

```{r}
#| label: run-round-1
lookup_def <- function(code, scheme) {
  tbl <- switch(scheme, ANZSCO = anzsco_defs, SOC2010 = soc_defs, ISCO08 = isco_defs)
  tbl[tbl$code == code, ]
}

rate_pair <- function(source_code, source_scheme, target_code) {
  src <- lookup_def(source_code, source_scheme)
  tgt <- lookup_def(target_code, "ISCO08")

  chat <- chat_openai(model = llm_model, echo = FALSE)
  rating <- chat$chat_structured(
    build_prompt(src$title, src$definition, tgt$title, tgt$definition),
    type = similarity_type
  )

  tibble(
    source_scheme, source_code, target_code,
    rating,
    model = llm_model, date = run_date
  )
}

ratings <- links_to_rate |>
  pmap(function(source_code, target_code, source_scheme) {
    rate_pair(source_code, source_scheme, target_code)
  }) |>
  list_rbind()

ratings
```

## Convert ratings to normalized weights

```{r}
#| label: normalize-weights
rating_score <- c(A = 1, B = 0.75, C = 0.5, D = 0.25, E = 0)

llm_weights <- ratings |>
  mutate(score = rating_score[rating]) |>
  group_by(source_scheme, source_code) |>
  mutate(weight = score / sum(score)) |>
  ungroup() |>
  select(source_scheme, source_code, target_code, rating, weight, model, date)

llm_weights
```

## What this produces

For each many-to-many source occupation, a set of weights across its linked ISCO-08 targets — generated by one documented prompt, one named model, on one recorded date. That's the whole point: the judgment call that would otherwise be an unrecorded line in someone's head (or an unexplained number in a script) is now a re-runnable artifact.

```{r}
#| label: save-output
write_csv(llm_weights, file.path(subset_dir, "llm_similarity_weights_prototype.csv"))
```

## Caveats (kept brief, per Ma, Huang & Haensch 2025)

- **One model, one run.** The paper found real variability across models (F1 41-70%) and low inter-model agreement (Cohen's κ < 60% between every pair). A different model, or the same model on a different day, could rate these pairs differently.
- **Not a replacement for review.** The paper's own framing is an assisted, human-in-the-loop workflow, not full automation. For a subset this small, a human can and should read every rating above directly rather than trust it unreviewed.
- **Not validated against anything.** There is no ground truth to check these weights against — this demonstrates that the judgment can be made explicit and reproducible, not that it's more (or less) accurate than any other method.
